Supplement 1: Different Response Distributions

### 2018-09-29 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
finished supplement improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2301 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 655ba80a0f84d600cb1a8ca78549694632c2781d

# boxplots with matplot
par(mfrow = c(1,4), mar = c(4,4,3,0.7))
matplot(t(s_df[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE,  main = "Distributional forest",
        xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_df, add = TRUE, col = "transparent")
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Distributional forest"), main = "Distributional forest",
#        ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,2.9,3,1.8))
matplot(t(s_g[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE, main = "Prespecified GAMLSS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_g, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,1.8,3,2.9))
matplot(t(s_gb[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE, main = "Boosted GAMLSS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_gb, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,0.7,3,4))
matplot(t(s_emos[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE,  main = "EMOS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_emos, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "EMOS"), main = "EMOS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
@
\caption{\label{fig:box_crpss_gaussian} CRPS skill score (higher = better)
for hurdle Gaussian and censored logistic distributions in reference to the
censored Gaussian distribution. Lines correspond to the 15 stations and are
aggregated in the boxplot. The y-axis range is limited
to $-10$\%, lower values for the two GAMLSS models are not displayed.}
\end{figure}

@@

\begin{figure}
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
# crps skill score by method (reference: EMOS)
means$crps_ss_method <- means$crps
means$crps_ss_method[means$method == "Distributional forest"] <- 
  1 - means$crps[means$method == "Distributional forest"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "Prespecified GAMLSS"] <- 
  1 - means$crps[means$method == "Prespecified GAMLSS"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "Boosted GAMLSS"] <- 
  1 - means$crps[means$method == "Boosted GAMLSS"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "EMOS"] <- 
  1 - means$crps[means$method == "EMOS"] / means$crps[means$method == "EMOS"]

@@

\begin{figure}
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7.2>>=

@@

As explained above, the parameter $\nu$ is modeled separately from $\mu$ and $\sigma$ 
in the GAMLSS framework which is why two separate models can be built. For the forest model
an alternative approach would be to build just one model employing the three-parametric
distribution family. However, testing both versions of the forest model in this application 
study shows that they both perform almost equally. Figure~\ref{fig:box_crpss_forests} shows
the CRPS skill score of the three-parametric forest model over the considered 15 observation
stations with the two-part forest model as reference. Even though this alternative version
of the forest model seems to perform better on average, according to the median this 
improvement is not even one percent, hence neglectably small. Therefore, the two forest 
versions can be stated to perform almost equally well.

\begin{figure}
\center
\setkeys{Gin}{width=0.5\textwidth}
<<compare_forests, fig=TRUE, results=hide, echo=FALSE, width=4, height=4>>=
load("crps_forests.rda")
crps_ss_forest_2p <- data.frame(1 - crps_forests[,"forest_2p"] / crps_forests[,"forest_h"])
names(crps_ss_forest_2p) <- "Three-parametric forest"
boxplot(crps_ss_forest_2p, ylab = "CRPS skill score", names = c("Three-parametric forest"))
abline(h = 0, col = pal[5], lwd = 2)
axis(1, 0:2, c("","Three-parametric forest", ""), las=1)
@
\caption{\label{fig:box_crpss_forests}CRPS skill score values of each of 
the 15 observation stations for the three-parametric forest model with 
the two-part forest model as reference model.}
\end{figure}

\newpage
%bibliographystyle{jss}
\bibliography{ref_supplement.bib}

>>>>

# boxplots with matplot
par(mfrow = c(1,4), mar = c(4,4,3,0.7))
matplot(t(s_df[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE,  main = "Distributional forest",
        xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_df, add = TRUE, col = "transparent")
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Distributional forest"), main = "Distributional forest",
#        ylim = c(-0.12,0.07), las = 2, ylab = "CRPS skill score")
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,2.9,3,1.8))
matplot(t(s_g[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE, main = "Prespecified GAMLSS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_g, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Prespecified GAMLSS"), main = "Prespecified GAMLSS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,1.8,3,2.9))
matplot(t(s_gb[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE, main = "Boosted GAMLSS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_gb, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "Boosted GAMLSS"), main = "Boosted GAMLSS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
par(mar = c(4,0.7,3,4))
matplot(t(s_emos[,]), type = "l", lwd = 2, 
        col = gray(0.5, alpha = 0.2),
        lty = 1, axes = FALSE,  main = "EMOS",
        xlab = "", ylab = "", xlim = c(0.5, 2.5),
        ylim = c(-0.10, 0.05))
boxplot(s_emos, add = TRUE, col = "transparent", yaxt = 'n')
abline(h = 0, col = pal[5], lwd = 2)
#boxplot(crps_ss_dist~distribution, data = means_sel, 
#        subset = (method == "EMOS"), main = "EMOS",
#        ylim = c(-0.12,0.07), las = 2)
#abline(h = 0, col = pal[5], lwd = 2)
@
\caption{\label{fig:box_crpss_gaussian} CRPS skill score (higher = better)
for hurdle Gaussian and censored logistic distributions in reference to the
censored Gaussian distribution (orange horizontal line). Gray lines correspond to the 15~stations and are
aggregated in the boxplot. The y-axis range is limited
to $-10$\%, lower values for the two GAMLSS models are not displayed.}
\end{figure}

\subsubsection*{FIXME: FURTHER GRAPHICS?}

\emph{The same data already shown in Figures~\ref{fig:lattice} and~\ref{fig:box_crpss_gaussian}
can be displayed in further ways. However, these do not really add much and hence should probably
be omitted.}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crpss_EMOS, fig=TRUE, echo=FALSE, height=3, width=7.2>>=
# crps skill score by method (reference: EMOS)
means$crps_ss_method <- means$crps
means$crps_ss_method[means$method == "Distributional forest"] <- 
  1 - means$crps[means$method == "Distributional forest"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "Prespecified GAMLSS"] <- 
  1 - means$crps[means$method == "Prespecified GAMLSS"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "Boosted GAMLSS"] <- 
  1 - means$crps[means$method == "Boosted GAMLSS"] / means$crps[means$method == "EMOS"]
means$crps_ss_method[means$method == "EMOS"] <- 
  1 - means$crps[means$method == "EMOS"] / means$crps[means$method == "EMOS"]

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.99\textwidth}
<<results_box_crps, fig=TRUE, echo=FALSE, height=3, width=7.2>>=

@@

\section{Discussion}

The comparison of the zero-censored Gaussian models from the main manuscript with the
alternative distributional specifications in this supplement reveals that a Gaussian hurdle
may lead to small improvements. However, especially for EMOS and the boosted GAMLSS model
the outcome is also considerably more variable. Therefore, overall there is no convincing
enough improvement over the zero-censored Gaussian model which is therefore the focus of
the main manuscript. Also, the distributional forest appears to be more robust against
distributional (mis-)specifications compared to the other models.

As a final consideration, we briefly remark that the distributional forest also allows
for a different hurdle model specification. Previously, we argued that for GAMLSS/EMOS
the log-likelihood of the hurdle model can be separated additively and hence the maximum
likelihood fit can be obtained by fitting two separate models for $\nu$ on one hand and
$\mu$ and $\sigma$ on the other. And therefore the same two-part approach was also adopted
for the distributional forests. However, an additional alternative is to fit a single
distributional forest based on the three-parameter log-likelihood from Equation~\ref{eq:hurdle}.
When all three parameters change simultaneously across the covariates, then this specification
can be expected to perform somewhat better. Whereas when different variables/effects drive
the parameters of the model, then the two-part approach implemented in this supplement
should perform somewhat better. In this particular case study, though, neither approach
performs clearly better as Figure~\ref{fig:box_crpss_forests} reveals. The figure depicts
CRPS skill scores for the 15~stations of the one-part hurdle models with the two-part
hurdle model as the reference. Clearly, both hurdle specifications perform very
similarly with skill scores being mostly below 1\%.

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.5\textwidth}
<<compare_forests, fig=TRUE, results=hide, echo=FALSE, width=4, height=4>>=
load("crps_forests.rda")
crps_ss_forest_2p <- data.frame(1 - crps_forests[,"forest_2p"] / crps_forests[,"forest_h"])
names(crps_ss_forest_2p) <- "Three-parametric forest"
boxplot(crps_ss_forest_2p, ylab = "CRPS skill score", names = c("Three-parametric forest"))
abline(h = 0, col = pal[5], lwd = 2)
axis(1, 0:2, c("","One-part vs. two-part hurdle forest", ""), las=1)
@
\caption{\label{fig:box_crpss_forests}CRPS skill score (higher = better) for the one-part
hurdle distributional forest in reference to the two part hurdle distributional forest
(orange horizontal line).
}
\end{figure}

\bibliography{ref_supplement.bib}
